### 2018-03-07 chapter2/ms.tex
Chapter 2 labels fixed

\section{Introduction}
\label{sec:intro}
The era of gravitational wave (GW) astronomy is fast approaching with the Advanced Laser Interferometer Gravitational Wave Observatory \citep[aLIGO,][]{LIGOMainRef} and advanced Virgo \citep[Virgo,][]{VirgoC7Burst} interferometers expected to make the first direct GW detections in the $2015-18$ time frame. These detectors will observe the GW signature resulting from the inspiral and merger of compact objects (e.g., neutron star-neutron star (NS-NS), solar mass black hole--black-hole (BH-BH), or NS-BH binaries). At design sensitivity, the ALIGO/AVIRGO network will be able to detect NS-NS mergers to an average distance of $\sim$ 200 Mpc \citep{Abadie+10} and with a typical localization region of $\sim$ 100 deg$^2$ \citep{Fairhurst2009,Nissanke+11,Aasi+14}. While this will be a remarkable achievement for both physics and astrophysics, the detection of an electromagnetic (EM) counterpart is essential to maximize the science obtained from a GW event. The identification of an EM counterpart will provide numerous benefits including: improved localization leading to host galaxy identification, determination of the distance and energy scales, identification of the progenitor local environment, and insight into the hydrodynamics of the merger \citep[see e.g.,][]{Sylvestre2003,Stubbs2008,Phinney2010,Stamatikos+09, Fong+10,MetzgerBerger12}. Furthermore, the detection of an EM counterpart will assist in breaking modeling degeneracies to obtain more accurate measurements of binary properties from the GW signal \citep{HughesHolz03}.

@@

To help provide a framework for these observations, and to more broadly investigate the coupled issues of kilonova detectability and contaminant rejection, we present here a systematic study of the feasibility of kilonova detections and assessing the effects of a population of contaminant sources during follow-up of a GW trigger using wide-field optical telescopes. This study will guide the development of observational strategies for wide-field searches that maximize the likelihood of detecting a GW optical counterpart. The paper is organized as follows: In Section~\ref{sec:contaminants} we describe the kilonova models utilized in this work and develop the list of contaminating transients considered in this work. In Section~\ref{sec:phase} we  describe observational features that may be used to isolate kilonovae from the population of potential contaminants. In Section~\ref{sec:MCsims} we present detailed Monte Carlo simulations of observations to investigate the detectability of kilonovae and contaminant rejection for various observational strategies. In Section~\ref{sec:diff}, we address issues that arise from the possible lack of a deep pre-existing template. We then discuss the effects of modifications to the standard kilonovae models on our results in Section~\ref{sec:altkilo}. In Section~\ref{sec:disc}, we asses the expected performance of current and future wide-field telescopes in the framework of kilonova detectability and contaminant rejection and develop strategies for GW follow-up. Conclusions are presented in Section~\ref{sec:conc}.

@@

\section{Kilonovae and Contaminants}
\label{sec:contaminants}

@@

\subsection{Kilonovae}
\label{sec:kilo}
As discussed in Section~\ref{sec:intro}, the merger of NS-NS or NS-BH binaries is expected to produce a faint, rapidly-evolving optical/NIR transient powered by the radioactive decay of {\em r}-process elements formed in the ejecta. From a wide range of simulations it has been shown that the ejecta have a typical mass range of $\Mej \sim 10^{-3}-10^{-1} \; M_{\odot}$ with velocities of $\beta_{\text{ej}} \sim 0.1-0.3$, which depend on the initial mass asymmetry in the binary \citep{Rosswog+99,Rosswog05,Bauswein+13a,Rosswog+13,PiranNakar13}. The ejecta are expected to be neutron-rich, with typical electron fractions of $Y_e \lesssim 0.1$ \citep{Metzger+10}.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{./figs/chapter2/ch2_f1.pdf}
\caption{Top: Kilonova spectra from \citet{BarnesKasen13} at $\sim$ 1 d and $\sim$ 1 week with ejecta properties of $\Mej =10^{-2}\;M_{\odot}$ and $\beta_{\text{ej}} = 0.2$. Bottom: Light curve for the same model parameters computed at a fiducial distance of 200 Mpc in the DECam {\em r}, {\em i}- and {\em z}-band filters. \vspace{10pt}}
\label{fig:kilosample}
\end{figure}

In this work, we construct kilonova light curves using the time-resolved spectra of \citet{BarnesKasen13}. They consider a grid of models with $\Mej = 10^{-3}\;, 10^{-2} \;, \text{ and } 10^{-1} \;M_{\odot}$ and with $\beta_{\text{ej}} = $ 0.1, 0.2, and 0.3. Unless otherwise noted we only consider the average ejecta properties of $\Mej =10^{-2}\;M_{\odot}$ and $\beta_{\text{ej}} = 0.2$ since we do not know the distribution of ejecta properties a priori. We compute light curves using the DECam filter transmission curves\footnote{\url{http://www.ctio.noao.edu/noao/sites/default/files/DECam/DECam_filters_transmission.txt}}\footnote{We note that the DECam transmission filters are fairly standard so these results are relevant for other instruments as well.} with the luminosity $L_{\phi}$, through a filter $\phi$, at a time t, being given by 
\begin{align}
\label{eq:LCint}
L_{\phi}(t) = \frac{\int d\lambda L_{\lambda}(t,\lambda) \phi(\lambda)}{\int d\lambda \phi(\lambda)}
\end{align}
\noindent where $L_{\lambda}(t,\lambda)$ is the model specific luminosity and $\phi(\lambda)$ is the filter transmission as a function of wavelength. A representative plot of the spectra at 1 day and 1 week, along with the resulting {\em r}-, {\em i}-, and {\em z}-band light curves are shown in Figure~\ref{fig:kilosample}.

\subsection{WD-NS/WD-BH Mergers}
\label{sec:WDmerge}
The tidal disruption of a WD by a NS or BH binary companion can produce an isotropic optical transient powered by the radioactive decay of ${}^{56}$Ni synthesized in the ejecta. The ejecta have an expected mass of $\Mej \approx 0.3-1.0 \; M_{\odot}$, velocity $\vej \approx (1-5) \times 10^4$ km s$^{-1}$, and $^{56}$Ni mass of $M_{\text{$^{56}$Ni}} \approx 10^{-4.5} - 10^{-2.5}\;M_{\odot}$ \citep{Metzger2012}. These properties are primarily determined by the mass of the WD, with more massive progenitors producing higher ejecta mass, velocities, and $^{56}$Ni masses. The resulting light curves have a typical peak timescale of $t_p \sim $ week and a peak bolometric luminosity of $L_p \sim 10^{39} - 10^{41.5} \text{ erg s}^{-1}$.

The two scenarios we consider here are the merger of: (1) a $0.6\;M_{\odot}$ C--O WD with a $1.4\;M_{\odot}$ NS, and (2) a $1.2\;M_{\odot}$ O--Ne WD with a $3\;M_{\odot}$ BH. The models of \citet{Metzger2012} only include bolometric light curves. To produce {\em i}- and {\em z}-band light curves we compute an effective temperature using the \citetalias{LP98} models with $\Mej \approx 0.5\;M_{\odot},\text{ and }\vej \approx 2.8\times10^4$ km s$^{-1}$ for scenario 1, and $\Mej \approx 1\;M_{\odot},\text{ and } \vej \approx 4.6\times10^4$ km s$^{-1}$ for scenario 2 \citep[see Figure 7 of][]{Metzger2012}. We integrate the resulting blackbody curve over the DECam filters using Equation~\ref{eq:LCint} and compare to the bolometric luminosity to generate bolometric corrections. For the WD-NS merger, we find $T_p \approx 6000$ K leading to $m_{\text{bol}} - m_{\text{i}} \approx -1.9$ mag and $m_{\text{bol}} - m_{\text{z}} \approx -2.2$ mag. The WD-BH merger calculation yields $T_p \approx 7850$ K leading to $m_{\text{bol}} - m_{\text{i}} \approx -2.2$ mag and $m_{\text{bol}} - m_{\text{z}} \approx -2.6$ mag. We assume that these corrections are constant in time. The bolometric corrections are applied to the light curves of \citet{Metzger2012} to generate the light curves used in this analysis (see Figure~\ref{fig:LCcont}).

@@

We use the models of \citet{Dabhra+10} to compute \emph{i} and \emph{z}-band light curves using the same methodology as in Section~\ref{sec:WDmerge}. The effective blackbody temperature has an assumed fiducial value of $T_p \approx 5800$ K from the models of \citet[see their Figure 3]{Dabhra+10}. The resulting bolometric corrections are therefore $m_{\text{bol}} - m_{\text{i}} \approx -1.9$ mag and $m_{\text{bol}} - m_{\text{z}} \approx -2.2$ mag. We assume that these corrections are constant in time and apply them to the bolometric light curves to generate $i$- and $z$-band light curves (Figure 2).

\subsection{White Dwarf Double Detonation}
\label{sec:ELDD}
Detonation of a sufficiently large accreted He envelope $(M_{\text{He}} \sim 0.2\; M_{\odot})$ on the surface of a C--O WD can generate shock waves which then propagate to the core leading to the total detonation of the WD. In simulations performed by \citet{Sim+12}, two possible detonation scenarios were considered. The first, edge-lit double detonation (ELDD), occurs when the initial detonation takes place at the surface of the WD. This produces a transient that is expected to reach a peak bolometric luminosity of $L_p \sim 10^{42}$ erg s$^{-1}$ at $t_p \sim 8$ d. The second scenario is the compression-shock double detonation (CSDD), where the core is compressed by the shock wave and C--O detonation occurs closer to the center of the WD. The transient is brighter than in the ELDD scenario with a peak bolometric luminosity of $L _p \sim 4\times10^{42}$ erg s$^{-1}$, and a correspondingly longer rise time ($t_p  \gtrsim 10$ d). In the CSDD scenario the transient is also longer-lived with no significant decline in brightness for $\sim 30$ d.

The slow evolution of the CSDD model light curve relative to kilonovae means that it can be easily removed from consideration as a contaminant on the basis of timescales alone. However, we still consider the ELDD scenario as the more rapid rise time and comparably faster decline means that it could serve as a potential contaminant. We make use of bolometric light curves from the simulations of \citet{Sim+12} to compute DECam {\em i}- and {\em z}-band light curves as in Section~\ref{sec:WDmerge}. The ELDD spectrum is approximated as a blackbody with a characteristic temperature of $T_p \approx 4800$ K using the model spectra of \citet[see their Figure 7]{Sim+12}. The bolometric corrections are found to be $m_{\text{bol}} - m_{\text{i}} \approx -1.7$  mag and $m_{\text{bol}} - m_{\text{z}} \approx -1.9$ mag. We assume that these corrections are constant in time and apply them to the bolometric light curves to generate $i$- and $z$-band light curves (Figure 2).

\subsection{Type .Ia Supernovae}
\label{sec:type.Ia}
Another possible WD-based transient is the detonation of an outer shell of helium on the surface of the WD. This can occur when the mass accretion rate from the donor is low enough that burning in the He shell becomes thermally unstable. The specific term ``type .Ia" was coined as a reference to He detonation in an AM Canum Venaticorum system (AM CVns) where the donor star is degenerate or semi-degenerate and accreting He rich material onto a C--O or O--Ne WD \citet{Bildsten+07,Shen+10}. The typical envelope mass is $M_{\text{env}} \sim \text{few}\times10^{-2}-10^{-1}\;M_{\odot}$ and the detonation ejects the bulk of the envelope at a typical velocity of $\vej \sim 10^4$ km s$^{-1}$. This leads to a transient with a typical peak time of $t_p \sim 2-10$ d and a peak bolometric luminosity of $L_p \sim 5\times10^{41} - 5\times 10^{42}$ erg s$^{-1}$ \citep{Shen+10}.

We produce type .Ia light curves from the time-resolved spectra computed from the models of \citet{Shen+10}. We consider the models $0.6+0.2\;M_{\odot}$, $0.6+0.3\;M_{\odot}$, $1.0+0.05\;M_{\odot}$, and $1.2+0.05\;M_{\odot}$, where the first number is the WD mass and the second number is the mass of the He envelope. We integrate the spectra over the DECam filter transmission curves as described in Section~\ref{sec:kilo} (Figure~\ref{fig:LCcont}).

\subsection{Pan-STARRS Fast Transients}
\label{sec:PSFast}
\citet{Drout+14} discovered ten rapidly evolving transients in the Pan-STARRS1 Medium Deep Survey. These events are characterized by a time above half-maximum of less than 12 days and have typical absolute magnitudes of $M \approx -16.5 \text{ to } -20$ mag. These objects have blue colors with {\em g} $-$ {\em r} $\lesssim$ --0.2 at peak \citep{Drout+14}. While the existing sample is small these events may end up being a significant contaminant in the fast cadence, wide-field GW searches investigated here. Therefore, we consider the {\em i}- and {\em z}-band light curves of three objects from \citet{Drout+14}: PS1-10ah, PS1-10bjp, and PS1-11qr. They possess sufficient pre-peak data for our analysis while providing a representative range of timescales and luminosities for the sample (Figure~\ref{fig:LCcont}).

\subsection{Type Ia Supernovae}
\label{sec:typeIa}
Type Ia supernova result from the thermonuclear explosion of a C--O WD. While these SNe have timescales much longer than those of kilonovae and other transients considered in this section, their high volumetric rate and large luminosity mean that a deep optical search for kilonovae will detect large numbers of Type Ia SNe. This makes them a particularly likely contaminant in such searches. We focus our analysis on the event SN2011fe discovered in M101 by the Palomar Transit Factory \citep{Nugent+11a} due to the high quality of available data. In addition to being one of the closest known Type Ia SNe \citep[$\mu = 29.04\pm0.19$,][]{ShappeeStanek11}, SN2011fe was discovered a mere 12 hours after explosion \citep{Nugent+11b} leading to coverage starting at --15 days relative to the {\em B}-band peak (R). In this work, we use the 32 spectral observations compiled by \citet{Pereira+13} to produce DECam {\em i}- and {\em z}-band light curves as a function of redshift, including K-corrections, using the process described in Section~\ref{sec:kilo} (Figure~\ref{fig:LCcont}).

\begin{figure}[h!]
\centering
\includegraphics[trim= 0 0 0 240 ,clip,width=0.6\textwidth]{./figs/chapter2/ch2_f2.pdf}
\caption{Top: {\em i}-band light curves of the contaminants considered in this work at a fiducial distance of 200 Mpc. Bottom: {\em z}-band light curves. See Sections~\ref{sec:WDmerge}~--~\ref{sec:typeIa} for details.}
\label{fig:LCcont}
\end{figure}

\subsection{Stellar Flares}
\label{sec:flarestar}
Stellar flares, primarily from M dwarfs, are another potential contaminant in fast cadence wide-field optical searches \citep{Becker+04,KulkarniRau06,Berger+13}.  These flares have characteristic timescales of minutes to hours \citep[see e.g.,][]{Berger+13} and temperature of $T_p \sim 10^4$ K, leading to an appearance as fast and blue transients.  As a result, M dwarf flares will only appear as a significant contaminant in kilonova searches that rely on single-epoch detection, primarily in $g$- and $r$-band in which the kilonova timescale is shorter than in the redder bands.  We return to this point in Section~\ref{sec:disc}.  The sky-projected M dwarf flare rate at the depths relevant for kilonova searches is large, $\sim 10$ deg$^{-2}$ day$^{-1}$ \citep{KulkarniRau06}.  On the other hand, as was demonstrated by \citet{Berger+13}, deep observations in the $i$- and $z$-band can easily reveal the underlying M dwarfs and hence reduce the potential for contamination from flares.

\subsection{AGN Variability}
\label{sec:agns}
We also consider short-duration optical variability from active galactic nuclei (AGNs). AGN typically show strong variability on month or year-long timescales, making AGN outbursts too long in duration to be confused with a kilonovae. However, recent work has detected rare cases of day-long  or shorter variability in the NIR. \citet{Genzel+03} observed an H-band flare in Sgr A$^{*}$ during which the source brightened by a factor of 5. This flare lasted $\sim$ 30 min, with a characteristic rise time of $\sim2-5$ min. Such short timescale flares can be ruled out as kilonova contaminants by requiring a detection on more than a single night. Flares with longer durations are also unlikely to be significant contaminants. \citet{Totani+05} performed {\em V}- and {\em i}-band observations of six AGN, selected on the basis of their variability, over a period of four days. They found typical variability of only $\sim1-5\%$ above their host galaxy luminosity. More recent work by \citet{Ruan+012} studied the variability of a sample of \fermi selected blazars. They found that these sources showed a typical rms variability of $\sim0.6$ mag on timescales of $\sim 10$ d. Therefore, AGN are unlikely to exhibit sufficiently large variability to be confused with a kilonova.

\subsection{Volumetric Rates}
\label{sec:rates}
An additional important factor for understanding the impact of potential contaminants on fast cadence, wide-field searches is their volumetric rates. We note that the rates for these types of objects are generally poorly understood due to the lack of strong observational constraints; the notable exception are Type Ia SNe. The volumetric rates of potential contaminants are compiled in Table~\ref{tab:rates}. No values are provided in the literature for the ELDD volumetric rate so we assign a fiducial value of 1\% of the Type Ia SN rate, consistent with the other rapid transients in this study. The volumetric rate for Type Ia SNe is known to be a function of redshift, and we use an approximate average volumetric rate in redshift bins of $\Delta z \sim 0.25$ out to $z_{\text{max}} \sim 0.75$. \citep[][and references therein]{Li+11,Graur+14}. These rates are listed in Table~\ref{tab:rates_Iae}.

>>>>

\section{Introduction}
\label{sec:ch2_intro}
The era of gravitational wave (GW) astronomy is fast approaching with the Advanced Laser Interferometer Gravitational Wave Observatory \citep[aLIGO,][]{LIGOMainRef} and advanced Virgo \citep[Virgo,][]{VirgoC7Burst} interferometers expected to make the first direct GW detections in the $2015-18$ time frame. These detectors will observe the GW signature resulting from the inspiral and merger of compact objects (e.g., neutron star-neutron star (NS-NS), solar mass black hole--black-hole (BH-BH), or NS-BH binaries). At design sensitivity, the ALIGO/AVIRGO network will be able to detect NS-NS mergers to an average distance of $\sim$ 200 Mpc \citep{Abadie+10} and with a typical localization region of $\sim$ 100 deg$^2$ \citep{Fairhurst2009,Nissanke+11,Aasi+14}. While this will be a remarkable achievement for both physics and astrophysics, the detection of an electromagnetic (EM) counterpart is essential to maximize the science obtained from a GW event. The identification of an EM counterpart will provide numerous benefits including: improved localization leading to host galaxy identification, determination of the distance and energy scales, identification of the progenitor local environment, and insight into the hydrodynamics of the merger \citep[see e.g.,][]{Sylvestre2003,Stubbs2008,Phinney2010,Stamatikos+09, Fong+10,MetzgerBerger12}. Furthermore, the detection of an EM counterpart will assist in breaking modeling degeneracies to obtain more accurate measurements of binary properties from the GW signal \citep{HughesHolz03}.

@@

To help provide a framework for these observations, and to more broadly investigate the coupled issues of kilonova detectability and contaminant rejection, we present here a systematic study of the feasibility of kilonova detections and assessing the effects of a population of contaminant sources during follow-up of a GW trigger using wide-field optical telescopes. This study will guide the development of observational strategies for wide-field searches that maximize the likelihood of detecting a GW optical counterpart. The paper is organized as follows: In Section~\ref{sec:ch2_contaminants} we describe the kilonova models utilized in this work and develop the list of contaminating transients considered in this work. In Section~\ref{sec:ch2_phase} we  describe observational features that may be used to isolate kilonovae from the population of potential contaminants. In Section~\ref{sec:ch2_MCsims} we present detailed Monte Carlo simulations of observations to investigate the detectability of kilonovae and contaminant rejection for various observational strategies. In Section~\ref{sec:ch2_diff}, we address issues that arise from the possible lack of a deep pre-existing template. We then discuss the effects of modifications to the standard kilonovae models on our results in Section~\ref{sec:ch2_altkilo}. In Section~\ref{sec:ch2_disc}, we asses the expected performance of current and future wide-field telescopes in the framework of kilonova detectability and contaminant rejection and develop strategies for GW follow-up. Conclusions are presented in Section~\ref{sec:ch2_conc}.

@@

\section{Kilonovae and Contaminants}
\label{sec:ch2_contaminants}

@@

\subsection{Kilonovae}
\label{sec:ch2_kilo}
As discussed in Section~\ref{sec:ch2_intro}, the merger of NS-NS or NS-BH binaries is expected to produce a faint, rapidly-evolving optical/NIR transient powered by the radioactive decay of {\em r}-process elements formed in the ejecta. From a wide range of simulations it has been shown that the ejecta have a typical mass range of $\Mej \sim 10^{-3}-10^{-1} \; M_{\odot}$ with velocities of $\beta_{\text{ej}} \sim 0.1-0.3$, which depend on the initial mass asymmetry in the binary \citep{Rosswog+99,Rosswog05,Bauswein+13a,Rosswog+13,PiranNakar13}. The ejecta are expected to be neutron-rich, with typical electron fractions of $Y_e \lesssim 0.1$ \citep{Metzger+10}.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{./figs/chapter2/ch2_f1.pdf}
\caption{Top: Kilonova spectra from \citet{BarnesKasen13} at $\sim$ 1 d and $\sim$ 1 week with ejecta properties of $\Mej =10^{-2}\;M_{\odot}$ and $\beta_{\text{ej}} = 0.2$. Bottom: Light curve for the same model parameters computed at a fiducial distance of 200 Mpc in the DECam {\em r}, {\em i}- and {\em z}-band filters. \vspace{10pt}}
\label{fig:ch2_kilosample}
\end{figure}

In this work, we construct kilonova light curves using the time-resolved spectra of \citet{BarnesKasen13}. They consider a grid of models with $\Mej = 10^{-3}\;, 10^{-2} \;, \text{ and } 10^{-1} \;M_{\odot}$ and with $\beta_{\text{ej}} = $ 0.1, 0.2, and 0.3. Unless otherwise noted we only consider the average ejecta properties of $\Mej =10^{-2}\;M_{\odot}$ and $\beta_{\text{ej}} = 0.2$ since we do not know the distribution of ejecta properties a priori. We compute light curves using the DECam filter transmission curves\footnote{\url{http://www.ctio.noao.edu/noao/sites/default/files/DECam/DECam_filters_transmission.txt}}\footnote{We note that the DECam transmission filters are fairly standard so these results are relevant for other instruments as well.} with the luminosity $L_{\phi}$, through a filter $\phi$, at a time t, being given by 
\begin{align}
\label{eq:LCint}
L_{\phi}(t) = \frac{\int d\lambda L_{\lambda}(t,\lambda) \phi(\lambda)}{\int d\lambda \phi(\lambda)}
\end{align}
\noindent where $L_{\lambda}(t,\lambda)$ is the model specific luminosity and $\phi(\lambda)$ is the filter transmission as a function of wavelength. A representative plot of the spectra at 1 day and 1 week, along with the resulting {\em r}-, {\em i}-, and {\em z}-band light curves are shown in Figure~\ref{fig:ch2_kilosample}.

\subsection{WD-NS/WD-BH Mergers}
\label{sec:ch2_WDmerge}
The tidal disruption of a WD by a NS or BH binary companion can produce an isotropic optical transient powered by the radioactive decay of ${}^{56}$Ni synthesized in the ejecta. The ejecta have an expected mass of $\Mej \approx 0.3-1.0 \; M_{\odot}$, velocity $\vej \approx (1-5) \times 10^4$ km s$^{-1}$, and $^{56}$Ni mass of $M_{\text{$^{56}$Ni}} \approx 10^{-4.5} - 10^{-2.5}\;M_{\odot}$ \citep{Metzger2012}. These properties are primarily determined by the mass of the WD, with more massive progenitors producing higher ejecta mass, velocities, and $^{56}$Ni masses. The resulting light curves have a typical peak timescale of $t_p \sim $ week and a peak bolometric luminosity of $L_p \sim 10^{39} - 10^{41.5} \text{ erg s}^{-1}$.

The two scenarios we consider here are the merger of: (1) a $0.6\;M_{\odot}$ C--O WD with a $1.4\;M_{\odot}$ NS, and (2) a $1.2\;M_{\odot}$ O--Ne WD with a $3\;M_{\odot}$ BH. The models of \citet{Metzger2012} only include bolometric light curves. To produce {\em i}- and {\em z}-band light curves we compute an effective temperature using the \citetalias{LP98} models with $\Mej \approx 0.5\;M_{\odot},\text{ and }\vej \approx 2.8\times10^4$ km s$^{-1}$ for scenario 1, and $\Mej \approx 1\;M_{\odot},\text{ and } \vej \approx 4.6\times10^4$ km s$^{-1}$ for scenario 2 \citep[see Figure 7 of][]{Metzger2012}. We integrate the resulting blackbody curve over the DECam filters using Equation~\ref{eq:LCint} and compare to the bolometric luminosity to generate bolometric corrections. For the WD-NS merger, we find $T_p \approx 6000$ K leading to $m_{\text{bol}} - m_{\text{i}} \approx -1.9$ mag and $m_{\text{bol}} - m_{\text{z}} \approx -2.2$ mag. The WD-BH merger calculation yields $T_p \approx 7850$ K leading to $m_{\text{bol}} - m_{\text{i}} \approx -2.2$ mag and $m_{\text{bol}} - m_{\text{z}} \approx -2.6$ mag. We assume that these corrections are constant in time. The bolometric corrections are applied to the light curves of \citet{Metzger2012} to generate the light curves used in this analysis (see Figure~\ref{fig:ch2_LCcont}).

@@

We use the models of \citet{Dabhra+10} to compute \emph{i} and \emph{z}-band light curves using the same methodology as in Section~\ref{sec:ch2_WDmerge}. The effective blackbody temperature has an assumed fiducial value of $T_p \approx 5800$ K from the models of \citet[see their Figure 3]{Dabhra+10}. The resulting bolometric corrections are therefore $m_{\text{bol}} - m_{\text{i}} \approx -1.9$ mag and $m_{\text{bol}} - m_{\text{z}} \approx -2.2$ mag. We assume that these corrections are constant in time and apply them to the bolometric light curves to generate $i$- and $z$-band light curves (Figure 2).

\subsection{White Dwarf Double Detonation}
\label{sec:ch2_ELDD}
Detonation of a sufficiently large accreted He envelope $(M_{\text{He}} \sim 0.2\; M_{\odot})$ on the surface of a C--O WD can generate shock waves which then propagate to the core leading to the total detonation of the WD. In simulations performed by \citet{Sim+12}, two possible detonation scenarios were considered. The first, edge-lit double detonation (ELDD), occurs when the initial detonation takes place at the surface of the WD. This produces a transient that is expected to reach a peak bolometric luminosity of $L_p \sim 10^{42}$ erg s$^{-1}$ at $t_p \sim 8$ d. The second scenario is the compression-shock double detonation (CSDD), where the core is compressed by the shock wave and C--O detonation occurs closer to the center of the WD. The transient is brighter than in the ELDD scenario with a peak bolometric luminosity of $L _p \sim 4\times10^{42}$ erg s$^{-1}$, and a correspondingly longer rise time ($t_p  \gtrsim 10$ d). In the CSDD scenario the transient is also longer-lived with no significant decline in brightness for $\sim 30$ d.

The slow evolution of the CSDD model light curve relative to kilonovae means that it can be easily removed from consideration as a contaminant on the basis of timescales alone. However, we still consider the ELDD scenario as the more rapid rise time and comparably faster decline means that it could serve as a potential contaminant. We make use of bolometric light curves from the simulations of \citet{Sim+12} to compute DECam {\em i}- and {\em z}-band light curves as in Section~\ref{sec:ch2_WDmerge}. The ELDD spectrum is approximated as a blackbody with a characteristic temperature of $T_p \approx 4800$ K using the model spectra of \citet[see their Figure 7]{Sim+12}. The bolometric corrections are found to be $m_{\text{bol}} - m_{\text{i}} \approx -1.7$  mag and $m_{\text{bol}} - m_{\text{z}} \approx -1.9$ mag. We assume that these corrections are constant in time and apply them to the bolometric light curves to generate $i$- and $z$-band light curves (Figure 2).

\subsection{Type .Ia Supernovae}
\label{sec:ch2_type.Ia}
Another possible WD-based transient is the detonation of an outer shell of helium on the surface of the WD. This can occur when the mass accretion rate from the donor is low enough that burning in the He shell becomes thermally unstable. The specific term ``type .Ia" was coined as a reference to He detonation in an AM Canum Venaticorum system (AM CVns) where the donor star is degenerate or semi-degenerate and accreting He rich material onto a C--O or O--Ne WD \citet{Bildsten+07,Shen+10}. The typical envelope mass is $M_{\text{env}} \sim \text{few}\times10^{-2}-10^{-1}\;M_{\odot}$ and the detonation ejects the bulk of the envelope at a typical velocity of $\vej \sim 10^4$ km s$^{-1}$. This leads to a transient with a typical peak time of $t_p \sim 2-10$ d and a peak bolometric luminosity of $L_p \sim 5\times10^{41} - 5\times 10^{42}$ erg s$^{-1}$ \citep{Shen+10}.

We produce type .Ia light curves from the time-resolved spectra computed from the models of \citet{Shen+10}. We consider the models $0.6+0.2\;M_{\odot}$, $0.6+0.3\;M_{\odot}$, $1.0+0.05\;M_{\odot}$, and $1.2+0.05\;M_{\odot}$, where the first number is the WD mass and the second number is the mass of the He envelope. We integrate the spectra over the DECam filter transmission curves as described in Section~\ref{sec:ch2_kilo} (Figure~\ref{fig:ch2_LCcont}).

\subsection{Pan-STARRS Fast Transients}
\label{sec:ch2_PSFast}
\citet{Drout+14} discovered ten rapidly evolving transients in the Pan-STARRS1 Medium Deep Survey. These events are characterized by a time above half-maximum of less than 12 days and have typical absolute magnitudes of $M \approx -16.5 \text{ to } -20$ mag. These objects have blue colors with {\em g} $-$ {\em r} $\lesssim$ --0.2 at peak \citep{Drout+14}. While the existing sample is small these events may end up being a significant contaminant in the fast cadence, wide-field GW searches investigated here. Therefore, we consider the {\em i}- and {\em z}-band light curves of three objects from \citet{Drout+14}: PS1-10ah, PS1-10bjp, and PS1-11qr. They possess sufficient pre-peak data for our analysis while providing a representative range of timescales and luminosities for the sample (Figure~\ref{fig:ch2_LCcont}).

\subsection{Type Ia Supernovae}
\label{sec:ch2_typeIa}
Type Ia supernova result from the thermonuclear explosion of a C--O WD. While these SNe have timescales much longer than those of kilonovae and other transients considered in this section, their high volumetric rate and large luminosity mean that a deep optical search for kilonovae will detect large numbers of Type Ia SNe. This makes them a particularly likely contaminant in such searches. We focus our analysis on the event SN2011fe discovered in M101 by the Palomar Transit Factory \citep{Nugent+11a} due to the high quality of available data. In addition to being one of the closest known Type Ia SNe \citep[$\mu = 29.04\pm0.19$,][]{ShappeeStanek11}, SN2011fe was discovered a mere 12 hours after explosion \citep{Nugent+11b} leading to coverage starting at --15 days relative to the {\em B}-band peak (R). In this work, we use the 32 spectral observations compiled by \citet{Pereira+13} to produce DECam {\em i}- and {\em z}-band light curves as a function of redshift, including K-corrections, using the process described in Section~\ref{sec:ch2_kilo} (Figure~\ref{fig:ch2_LCcont}).

\begin{figure}[h!]
\centering
\includegraphics[trim= 0 0 0 240 ,clip,width=0.6\textwidth]{./figs/chapter2/ch2_f2.pdf}
\caption{Top: {\em i}-band light curves of the contaminants considered in this work at a fiducial distance of 200 Mpc. Bottom: {\em z}-band light curves. See Sections~\ref{sec:ch2_WDmerge}~--~\ref{sec:ch2_typeIa} for details.}
\label{fig:ch2_LCcont}
\end{figure}

\subsection{Stellar Flares}
\label{sec:ch2_flarestar}
Stellar flares, primarily from M dwarfs, are another potential contaminant in fast cadence wide-field optical searches \citep{Becker+04,KulkarniRau06,Berger+13}.  These flares have characteristic timescales of minutes to hours \citep[see e.g.,][]{Berger+13} and temperature of $T_p \sim 10^4$ K, leading to an appearance as fast and blue transients.  As a result, M dwarf flares will only appear as a significant contaminant in kilonova searches that rely on single-epoch detection, primarily in $g$- and $r$-band in which the kilonova timescale is shorter than in the redder bands.  We return to this point in Section~\ref{sec:ch2_disc}.  The sky-projected M dwarf flare rate at the depths relevant for kilonova searches is large, $\sim 10$ deg$^{-2}$ day$^{-1}$ \citep{KulkarniRau06}.  On the other hand, as was demonstrated by \citet{Berger+13}, deep observations in the $i$- and $z$-band can easily reveal the underlying M dwarfs and hence reduce the potential for contamination from flares.

\subsection{AGN Variability}
\label{sec:ch2_agns}
We also consider short-duration optical variability from active galactic nuclei (AGNs). AGN typically show strong variability on month or year-long timescales, making AGN outbursts too long in duration to be confused with a kilonovae. However, recent work has detected rare cases of day-long  or shorter variability in the NIR. \citet{Genzel+03} observed an H-band flare in Sgr A$^{*}$ during which the source brightened by a factor of 5. This flare lasted $\sim$ 30 min, with a characteristic rise time of $\sim2-5$ min. Such short timescale flares can be ruled out as kilonova contaminants by requiring a detection on more than a single night. Flares with longer durations are also unlikely to be significant contaminants. \citet{Totani+05} performed {\em V}- and {\em i}-band observations of six AGN, selected on the basis of their variability, over a period of four days. They found typical variability of only $\sim1-5\%$ above their host galaxy luminosity. More recent work by \citet{Ruan+012} studied the variability of a sample of \fermi selected blazars. They found that these sources showed a typical rms variability of $\sim0.6$ mag on timescales of $\sim 10$ d. Therefore, AGN are unlikely to exhibit sufficiently large variability to be confused with a kilonova.

\subsection{Volumetric Rates}
\label{sec:ch2_rates}
An additional important factor for understanding the impact of potential contaminants on fast cadence, wide-field searches is their volumetric rates. We note that the rates for these types of objects are generally poorly understood due to the lack of strong observational constraints; the notable exception are Type Ia SNe. The volumetric rates of potential contaminants are compiled in Table~\ref{tab:ch2_rates}. No values are provided in the literature for the ELDD volumetric rate so we assign a fiducial value of 1\% of the Type Ia SN rate, consistent with the other rapid transients in this study. The volumetric rate for Type Ia SNe is known to be a function of redshift, and we use an approximate average volumetric rate in redshift bins of $\Delta z \sim 0.25$ out to $z_{\text{max}} \sim 0.75$. \citep[][and references therein]{Li+11,Graur+14}. These rates are listed in Table~\ref{tab:ch2_rates_Iae}.
